---
title: "Getting started with mmCTBN"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with mmCTBN}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

# Overview

`mmCTBN` ("multimorbidity continuous-time Bayesian networks") is a unified
framework for fitting CTBN models to longitudinal Electronic Health Record
data. It supports **five interchangeable estimation backends** under one
common interface (`ctbn_fit(..., method = ...)`):

  * **`map`** — fast regularised Poisson MAP with four prior families.
  * **`stan`** — full Bayesian MCMC.
  * **`classic`** — Nodelman et al. (2002) closed-form MLE per parent
    configuration (no covariates).
  * **`fctbn`** — Faruqui et al. (2021) Functional CTBN: Poisson
    regression with adaptive group lasso and multiplicative parent
    effects, optimised with FISTA + Gaussian-mixture early stop.
  * **`cph`** — Guillamet et al. (2025) CTBN-CPH: classical baseline
    intensities modulated per directed edge by Cox proportional-hazards
    regressions, yielding patient-specific rates.

The package is designed for high-dimensional multimorbidity studies
where one cares not just about pairwise associations but also about
**synergistic 2-, 3-, 4-, and 5-way interactions** between conditions
(via `method = "map"` or `"stan"`).

```{r}
library(mmCTBN)
library(data.table)
```


# 1. Simulating data

Generate a 5-node random network with 2-way and 3-way interactions:

```{r}
set.seed(2026)
nw <- make_random_network(n_nodes = 5, seed = 2026)
DT <- simulate_ctbn_data(
  network            = nw,
  n_patients         = 1000,
  t_horizon          = 10,
  dt_step            = 1,
  interaction_order  = 3,         # 1 = mains only, ..., 5 = up to 5-way
  parallel           = TRUE,
  n_cores            = 4)
head(DT)
```

For the canonical 10-condition multimorbidity benchmark used in the
original paper, use:

```{r}
nw10 <- make_default_network()
DT10 <- simulate_ctbn_data(network = nw10, n_patients = 2000,
                           interaction_order = 2)
```


# 2. Fast MAP fit

```{r}
fit_map <- ctbn_fit(
  DT,
  method            = "map",
  prior             = "spike_slab",
  max_order         = 2,
  fixed_covs        = c("age", "sex_male"),
  time_varying_covs = c("smk_current", "smk_former"),
  parallel          = TRUE)

summary(fit_map)
coef(fit_map, type = "beta")
plot(fit_map, type = "network")
plot(fit_map, type = "heatmap")
```


# 3. Full Bayesian fit

When you need credible intervals, posterior predictive checks, or full
uncertainty quantification, switch backends without changing any other
code:

```{r}
fit_stan <- ctbn_fit(
  DT,
  method            = "stan",
  prior             = "horseshoe",
  max_order         = 2,
  fixed_covs        = c("age", "sex_male"),
  time_varying_covs = c("smk_current", "smk_former"),
  iter              = 2000,
  chains            = 4,
  cores             = 4)
```


# 4. Inference

Cumulative incidence at a 5-year horizon for any target:

```{r}
patient0 <- DT[eid == DT$eid[1]]
F5 <- compute_F_m(fit_map, target = "C1", patient_dt = patient0, tau = 5)
```

Synergistic interaction decomposition:

```{r}
syn <- compute_interaction_effects(fit_map, DT, tau = 5,
                                    pip_thresh = 0.5)
plot_synergy_forest(syn, top_n = 15)
```


# 5. Cross-validation

```{r}
fit_fns <- list(
  map_ss = function(dt, ...)
    ctbn_map(dt, prior = "spike_slab", max_order = 2, parallel = FALSE,
             verbose = FALSE, ...),
  map_hs = function(dt, ...)
    ctbn_map(dt, prior = "horseshoe", max_order = 2, parallel = FALSE,
             verbose = FALSE, ...))

future::plan(future::multisession, workers = 4)
cv <- ctbn_cv(
  DT_wide   = DT,
  fit_fns   = fit_fns,
  k_folds   = 3,
  eval_times = c(2, 5, 8),
  fixed_covs = c("age", "sex_male"))
summarise_cv(cv)
plot_tdauc(cv)
```


# 6. Recovery / selection metrics (simulation studies)

If the true network is known (e.g. from `make_random_network`), recovery
quality can be evaluated with `compute_recovery_metrics()` and
`compute_selection_metrics()`. See `?compute_recovery_metrics` for
details.


# 7. Classical CTBN (Nodelman 2002)

The original CTBN has no covariate channel and estimates rates by
closed-form MLE per parent configuration. Useful as a sanity check and
for small problems with strong topology priors.

```{r}
fit_classic <- ctbn_fit(
  DT,
  method      = "classic",
  parents     = list(),       # default: every other condition is a parent
  max_parents = 3,
  verbose     = FALSE)
summary(fit_classic)
plot(fit_classic, type = "network")
```


# 8. Functional CTBN (Faruqui 2021)

Poisson regression with adaptive group lasso and multiplicative parent
effects. Each (parent, target) edge is a *group* in the FISTA solver;
sparsity comes from group-soft-thresholding. The GMM-based early stop
snaps near-zero groups to exactly zero after a warmup window, saving
many iterations.

```{r}
fit_fctbn <- ctbn_fit(
  DT,
  method            = "fctbn",
  fixed_covs        = c("age", "sex_male"),
  time_varying_covs = c("smk_current", "smk_former"),
  lambda            = 1e-2,    # group-lasso strength
  max_iter          = 20000,
  gmm_stop          = TRUE,
  gmm_warmup        = 3000)
summary(fit_fctbn)
plot(fit_fctbn, type = "heatmap")
```


# 9. CTBN-CPH (Guillamet 2025)

Hybrid two-stage model. First, baseline intensities $\lambda^0_{m|u}$
are estimated by the classical CTBN. Second, for every directed edge
$j \to m$ a Cox proportional-hazards regression on patient covariates
yields hazard ratios $\exp(\sum_l \beta_l c_l)$ which modulate the
baseline rate at inference time. Covariates with Cox p-values above
`alpha` are gated out.

```{r}
fit_cph <- ctbn_fit(
  DT,
  method            = "cph",
  fixed_covs        = c("age", "sex_male"),
  time_varying_covs = c("smk_current", "smk_former"),
  max_parents       = 3,
  alpha             = 0.05)
summary(fit_cph)
# Per-edge Cox coefficients
str(fit_cph$beta_cov, max.level = 2)
```


# 10. Cross-validation across methods

The CV harness works uniformly across all five backends: just hand
`ctbn_cv()` a list of fitters.

```{r}
fit_fns <- list(
  map     = function(dt, ...) ctbn_map    (dt, prior = "spike_slab",
                                            max_order = 2, ...),
  classic = function(dt, ...) ctbn_classic(dt, max_parents = 2, ...),
  fctbn   = function(dt, ...) ctbn_fctbn  (dt, fixed_covs = c("age"),
                                            lambda = 1e-2,
                                            max_iter = 5000, ...),
  cph     = function(dt, ...) ctbn_cph    (dt, fixed_covs = c("age"),
                                            max_parents = 2, ...))
future::plan(future::multisession, workers = 4)
cv <- ctbn_cv(DT, fit_fns = fit_fns, k_folds = 3,
              eval_times = c(2, 5, 8))
plot_tdauc(cv)
```
